---
title: "Using the wizard function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the wizard function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(BitsnBobs)
library(rlang)
```

Copyright (c) 2026 Mark Eisler

## 1. Introduction
The `wizard()` function extracts and sorts unique values of a selected column from a data frame, and optionally pastes into a character string.

For our examples, we will use the `mtcars` data.

## 2. Using a symbol argument
`wizard(data, col, .collapse = NULL)` uses a symbol for its `<data-masking>` argument `col`.

```{r}
mtcars |> wizard(cyl)
```
This shows sorted unique values of the `cyl` column of `mtcars`.

If there is a potentially confounding variable of the same name in the global environment,

```{r}
(cyl <- c(100:110))
```
...nevertheless, `<data-masking>` ensures that the `col` argument of `wizard()` correctly refers to the `cyl` column of the `mtcars` data.

```{r}
mtcars |> wizard(cyl)
```
But say there’s a similar variable `Cyl` in the global environment,

```{r}
Cyl <- "Bad Luck!"
```
...and we confuse `cyl` with `Cyl`, we obtain an unexpected and puzzling result: -

```{r}
mtcars |> wizard(Cyl)
```
## 3. Using the `.data` pronoun
Using the `.data` pronoun^[[**rlang** .data and .env pronouns](https://rlang.r-lib.org/reference/dot-data.html)] from the **rlang** package
circumvents this problem by throwing a meaningful error message: -
```{r}
mtcars |> wizard(.data$cyl)

try(mtcars |> wizard(.data$Cyl))
```
## 4. Symbolise and inject
The col argument might be saved as a character variable, for example: -

```{r}
(col_var <- "cyl")
```
...which needs converting to a symbol and injecting^[[**rlang** symbolise and inject](https://rlang.r-lib.org/reference/topic-metaprogramming.html#symbolise-and-inject)] using the injection operator `!!` from the **rlang** package: -

```{r}
mtcars |> wizard(!!sym(col_var))
```
Again, if we confuse `cyl` with `Cyl`, we obtain an unexpected and confusing result: -

```{r}
(col_var <- "Cyl")

mtcars |> wizard(!!sym(col_var))
```

Using `data_sym()` from **rlang** circumvents this problem by prefacing the symbol with the data pronoun: -

```{r}
data_sym(col_var)

try(mtcars |> wizard(!!data_sym(col_var)))
```

## 5. Examples using `lapply()`
(We could also use `purrr::map()`, but it has no particular advantage here so **base** `lapply()` is preferable.)

```{r}
mtcars |> lapply(wizard, data = mtcars)
```

Here we use `setNames()` to obtain an eponymously named character vector of the first four `mtcar` names: -

```{r}
(carnames <- setNames(nm = names(mtcars))[1:4])
```

...and use it in three equivalent versions where `lapply()` uses the `.data pronoun` ,`data_sym()` or `data_syms()` in an anonymous function: -

```{r}
carnames |> lapply(\(x) wizard(mtcars, .data[[x]]))

carnames |> lapply(\(x) wizard(mtcars, !!data_sym(x)))

carnames |> data_syms() |> lapply(\(x) wizard(mtcars, !!x))
```

If one of the names is incorrect: -
```{r}
carnames["cyl"] <- "Cyl"
carnames
```

...these all provide meaningful error messages as previously shown: -

```{r}
try(carnames |> lapply(\(x) wizard(mtcars, .data[[x]])))

try(carnames |> lapply(\(x) wizard(mtcars, !!data_sym(x))))

try(carnames |> data_syms() |> lapply(\(x) wizard(mtcars, !!x)))
```

Without the `.data` pronoun, the error might be easily overlooked: -

```{r}
carnames |> lapply(\(x) wizard(mtcars, !!sym(x)))

carnames |> syms() |> lapply(\(x) wizard(mtcars, !!x))
```

## 6. References
1. [**rlang** .data and .env pronouns](https://rlang.r-lib.org/reference/dot-data.html)
2. [**rlang** symbolise and inject](https://rlang.r-lib.org/reference/topic-metaprogramming.html#symbolise-and-inject)

